---
title: "Economic concepts"
author: "Luis Francisco Gomez Lopez"
date: 2020-07-13
output:
  html_document:
    code_folding: hide
    df_print: paged
    highlight: tango
    number_sections: yes
    theme: flatly
    toc: yes
    toc_depth: 4
bibliography: economic_concepts.bib
link-citations: yes
---

<style>

body {
text-align: justify}

</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo      = TRUE, 
                      warning   = FALSE, 
                      message   = FALSE,
                      fig.align = "center")
```

```{r libraries}

library(tidyverse)
library(wbstats)
library(plotly)
```

# Gross Domestic Product (GDP)

We can initially approach the concept of __GDP__ by explaining the terms __"Product"__, __"Domestic"__ and __"Gross"__ separately [@lequiller_understanding_2014, Chapter 1]:

- __Product:__ refers to what we are trying to measure, which is the production of goods and services, with no double counting, in a given period, $t$ and carried out by:

    + Profit-making enterprises (economists use the alternative term _firms_)
    
    + Non-profit institutions
    
    + Goverment bodies
    
    + Households

- __Domestic__: indicates that the production to be taken into account is the one that is carried within a certain geographical territory, $s$, clearly delimited. 

    + In that sense if we use the concept of of __GDP__ we don't care who carried out the production in $t$ but only where the production was carried out in $t$ 
    
- __Gross__: it means that depreciation is not deducted (in the field of economics it is called consumption of fixed capital). 

    + In other words, the decrease in the value of the assets due to physical deterioration, foreseeable wear or accidental damage is not deducted.
    
    + Why depreciation is not deducted?
    
        + It is a variable difficult to measure and the methods used to measure it tend to differ between countries creating problems for international comparisons
        
## Data about __GDP__ using  Colombia as an example

Initially the units  in which __GDP__ is measure is in monetary units of a specific currency, $c$. Therefore $GDP_{s}^{c}(t)$ means the $GDP$ of territory $s$ in a given period $t$. To make the discussion less abstract we present a plot of $GDP$ for Colombia, $s = COL$, expressed in Colombian pesos, $c = COP$, for the years 1960 to 2019, $t = 1960, \ldots, 2019$: 

```{r, fig.width = 7, fig.height = 8}

# Clean data
gdp_colombia <- wbstats::wb(country   = "COL",
                             indicator = "NY.GDP.MKTP.CN",
                             startdate = 1960,
                             enddate   = 2019) %>%
    tibble::as_tibble() %>%
    dplyr::select(date, value) %>%
    dplyr::mutate(date       = as.double(date),
                  label_text = stringr::str_glue('Year: {date}
                                                  GDP: {value %>% scales::dollar()}'))

# Plot

static_plot <- gdp_colombia %>%
    # Data
    ggplot2::ggplot(aes(x = date, y =value)) +
    # Geoms
    ggplot2::geom_point(aes(text = label_text),
                        shape = 21,
                        color = "black",
                        fill  = "red") +
    ggplot2::geom_line(linetype = "dashed") +
    # Scales
    scale_x_continuous(breaks = c(1960:2019)) +
    scale_y_continuous(breaks = seq(from = 0, to = 1.10e15, by = 1e14),
                       labels = scales::number_format(scale = 1/1e12, suffix = "B")) +
    labs(x = "Year",
         y = "GDP in current local currency [B = Billion in long scale (10^12)]",
         title = "GDP of Colombia: 1960-2019") +
    # Themes
    theme(panel.border      = element_rect(fill = NA, color = "black"),
          plot.background   = element_rect(fill = "#f3fcfc"),
          panel.background  = element_rect(fill = "#f3f7fc"),
          legend.background = element_rect(fill = "#f3fcfc"),
          plot.title        = element_text(face = "bold"),
          axis.text.x=element_text(angle = -90, vjust = 0.5),
          axis.title        = element_text(face = "bold"),
          legend.title      = element_text(face = "bold"),
          axis.text         = element_text(face = "bold"))

    # Interactivity
    static_plot %>%
      plotly::ggplotly(tooltip = "text")

```

## A simple economy to explain the measurement of production [@blanchard_macroeconomics_2017]

- In a certain geographical territory $s$ and in a given period $t$ there exist $2$ profit-making enterprises:

    + __Steel enterprise__ 
    
    + __Car enterprise__
    
        + The __Steel enterprise__ sells to the __Car enterprise__ steel. Then the __Car enterprise__ uses steel to produce cars an sell them to households located inside or outside the geographical territory $s$ 
        
        + In that sense the economy uses steel, machinery and labor, known as inputs of production in the field of economics , to produce a final good represented by cars where the production is divided between the owners of the $2$ enterprises who perceive an income represented by profits and the workers who perceive an income represented by wages.
        
- In this simple economy we do not have non-profit institutions and a government that imposes taxes to offer services and goods like public education in $t$ because the only goods produced are steel and cars. 

- We also assume that production doesn't accumulate in $t$. Therefore steel is used entirely to produce cars and all the cars are consumed by households located inside or outside the geographical territory $s$. Thus, the profit-making enterprises don't accumulate inventories and distribute all the profits to the owners of the $2$ enterprises in $t$.

- Finally we assume that households inside the geographical territory $s$ only consume cars produced inside the geographical territory $s$ and not cars produced outside the geographical territory $s$ in $t$. Also, the $2$ profit-making enterprises don't buy machines produce outside the geographical territory $s$ to produce steel or cars in $t$.

- We can have a more realistic economy that have:

    + Profit-making enterprises that acummulate inventories or don't distribute all the profits to the owners
    
    + Profit-making enterprises that not only produce but pollute the air or the environment
    
    + Non-profit institutions that by law are requiere not to distribute profits
    
    + Households located inside or outside the geographical territory $s$ that save part of their income and don't consume everything or that engage in illegal activities like robberies
    
    + Households inside the geographical territory $s$ that have children and require and education system
    
    + Households inside the geographical territory $s$ that consume goods and services produced outside the geographical territory $s$ 
    
    + A government that imposes taxes to offer services like justice and goods like public education and deliver subsidies
    
And many other aspects not included in this short list but the idea is to explain in a simple way the measurement of __GDP__.

- The above simple economy in a certain geographical territory $s$ and in a given period $t$ can be represented in the following way using specific values to be less abstract and expressing every item in specific currency, $c$, like Colombia pesos (COP):

    + __Steel enterprise__
    
        + Steel sales to __Car enterprise__: $100 \text{ COP}$
        
        + Expenses:
        
            + Wages: $80 \text{ COP}$
            
        + Profits: $20 \text{ COP}$
        
    + __Car enterprise__
    
        + Revenue from sales of cars: $200 \text{ COP}$
        
        + Expenses:
        
            + Wages: $70$ COP
            
            + Steel purchases to __Steel enterprise__: $100$ COP
            
        + Profits: $30$ COP
        
### Measuring production and the double counting problem

- If you add in monetary terms the production of both companies you get a total of: Production __Steel enterprise__ $+$ Production __Car enterprise__ $=$ Steel sales to __Car enterprise__ $+$ Revenue from sales of cars $= 100 \text{ COP} + 200  \text{ COP} = 300 \text{ COP}$ 

- If the production of the __Steel enterprise__ and the __Car enterprise__ is added, the value of steel is being added two times.

- It is necessary to eliminate at some stage of the production process the value of steel in our example

### Three equivalent ways to measure production and avoid the double counting problem 

#### __GDP__ as the sum of value added

- __GDP__ is the sum of value added in a certain geographical territory $s$ during a given period $t$ expressed in a local currency $c$.

- In the field of economics the value added is the value that is added at each stage of production. It is defined as the difference between the __Production in expressed in a Monetary Terms (PMT)__ and the __Consumption of Intermediate Goods (CIG)__.

    + The __PMT__ is simply the production expressed using a local currency $c$ 

    + The __CIG__ is the monetary value expressed using a local currency $c$ of inputs that are completely transformed and depleted in the production process and that are used to produce other products.
    
        + Example of inputs that are not part of the __CIG__:
        
            + __Wages paid by a profit-making enterprise to its workers__: labor can be used for several periods and although its value is affected in the periods close to the age of retirement of individuals, is not fully consumed in the production process.
            
            + __Assets that belong to a profit-making enterprise and depreciation (consumption of fixed capital)__: assets are durable goods that can be used for several periods and although their value is affected by physical deterioration, foreseeable wear and accidental damage it is important to remember that depreciation (consumption of fixed capital) is included and not deducted in the measurement of __GDP__.
            
- Measuring __GDP__ as the sum of value added using our simple economy:
            
    + __Steel enterprise__
    
        + __PMT__: $100 \text{ COP}$
        
        + __CIG__: $0 \text{ COP}$ (Wages are not part of __CIG__)
        
        + __Value added__: $100 \text{ COP} - 0 \text{ COP } = 100  \text{ COP}$
        
    + __Car enterprise__
    
        + __PMT__: $200 \text{ COP}$
        
        + __CIG__: $100 \text{ COP}$ (Wages are not part of __CIG__)
        
        + __Value added__: $200 \text{ COP} - 100 \text{ COP } = 100  \text{ COP}$
        
    + __GDP__
    
        + __Total Value Added__ = __Value added Steel enterprise__ $+$ __Value added Car enterprise__ $= 100 \text{ COP} + 100 \text{ COP}  = 200 \text{ COP} =$ __GDP__
        
#### __GDP__ as the sum of incomes

- __GDP__ is the sum of the incomes perceived by individuals in a certain geographical territory $s$ during a given period $t$ expressed in a local currency $c$.

- In that sense the __GDP__ can me measure as the sum of the different incomes that individuals perceive like profits and wages

- Measuring __GDP__ as the sum of incomes using our simple economy:

    + __Steel enterprise__
    
        + __Workers income__: $80 \text{ COP}$
        
        + __Owners income__: $20 \text{ COP}$
        
    + __Car enterprise__
    
        + __Workers income__: $70 \text{ COP}$
        
        + __Owners income__: $30 \text{ COP}$
        
    + __GDP__
    
        + __Total Income__ = __Total workers income__ $+$ __Total owners income__ $= (80 \text{ COP} + 20 \text{ COP}) +  (20 \text{ COP} + 30 \text{ COP}) = 200 \text{ COP} =$ __GDP__
        
#### __GDP__ as the value and uses of final goods and services

- __GDP__ is the value of all the final goods and services produced in a certain geographical territory $s$ during a given period $t$ expressed in a local currency $c$ and classified according to their use.

- We can classify the uses of final production in the following items which corresponds only to goods and services produced __inside__ a certain geographical territory $s$ during a given period $t$ expressed in a local currency $c$:

    + __Households and non-profit institutions final consumption expenditure__: $C_s^c(t)$
    
        + It simply indicates that part of the production is consumed by households which includes the expenditures of nonprofit institutions serving households.
        
            + There is an exception where purchases of dwellings by households are excluded from $C_s^c(t)$   
    
    +  __Gross capital formation__: $I_s^c(t)$
    
        + It simply refers to additions of fixed assets to the economy plus net changes in the level of inventories.  
        
            + In this item purchases of dwellings by households are included becauses dwellings are considered additions of fixed assets to the economy.
            
            + Inventories refers to stocks of goods held by enterprises to meet temporary or unexpected fluctuations in production or sales and work in progress (in our simple economy "work in progress" will be for example the case of a car that it is not totally finished by __Car enterprise__ at the end of period $t$).
            
    + __General government final consumption expenditure__: $G_s^c(t)$
    
        + It simply refers to most of the government's current expenses for purchases of goods and services.
        
    + __Exports of goods and services__: $X_s^c(t)$
    
        + It simply refers to all goods and services provided to the rest of the world that is outside $s$.
        
- In that sense $GDP_s^c(t) = C_s^c(t) + I_s^c(t) + G_s^c(t) + X_s^c(t)$

- Measuring __GDP__ as the as the value and uses of final goods and services:

    + $C_s^c(t) + X_s^c(t) = 200 \text{ COP}$ (It corresponds to the Revenue from sales of cars that are consumed by households or exported)
    
    + $I_s^c(t) = 0 \text{ COP}$
    
    + $G_s^c(t) = 0 \text{ COP}$
    
    + $GDP_s^c(t) = 200 \text{ COP} + 0 \text{ COP} + 0 \text{ COP} = 200 \text{ COP}$
    
##### The macroeconomic identity

- In all the macroeconomic textbooks and data published by statistical organizations it is pointed out that $GDP = C + I + G + X - IM$ where $IM$ refers to the imports received from the rest of the world in relation to a certain geographical territory $s$.

- This is consistent with $GDP_s^c(t) = C_s^c(t) + I_s^c(t) + G_s^c(t) + X_s^c(t)$ if we rewrite this expression as:

$$\begin{split}
   GDP_s^c(t) & = C_s^c(t) + I_s^c(t) + G_s^c(t) + X_s^c(t) \\
   & = C_s^c(t) + C_{rw}^c(t) + I_s^c(t) + I_{rw}^c(t) + G_s^c(t) + G_{rw}^c(t) + X_s^c(t) - C_{rw}^c(t) - I_{rw}^c(t) - G_{rw}^c(t) \\
   & = C_s^c(t) + C_{rw}^c(t) + I_s^c(t) + I_{rw}^c(t) + G_s^c(t) + G_{rw}^c(t) + X_s^c(t) - (C_{rw}^c(t) + I_{rw}^c(t) + G_{rw}^c(t)) \\
   & = C^c(t) + I^c(t) + G^c(t) + X^c(t) - IM^c(t) 
   \end{split}$$

- Where: 

    + $C_{rw}^c(t)$ is the __Households and non-profit institutions final consumption expenditure__ with goods and services produced in the rest of the world, $rw$.
    
    + $C^c(t) = C_s^c(t) + C_{rw}^c(t)$ is the _ Total Households and non-profit institutions final consumption expenditure__ 
    
    + $I_{rw}^c(t)$ is the __Gross capital formation__ with goods and services produced in the rest of the world, $rw$.
    
    + $I^c(t) = I_s^c(t) + I_{rw}^c(t)$ is the __Total Gross capital formation__
    
    + $G_{rw}^c(t)$ is the __General government final consumption expenditure__ with goods and services produced in the rest of the world, $rw$.
    
    + $G^c(t) = G_s^c(t) + G_{rw}^c(t)$ is the __Total General government final consumption expenditure__
    
    + $IM^c(t) = C_{rw}^c(t) + I_{rw}^c(t) + G_{rw}^c(t)$
    
## From a simple economy with artificial data to real data

Using data with annual periodicity provided by the Departamento Nacional de Estadística (DANE) from Colombia expressed in Colombian Pesos (COP) for the years $2005-2019$ we show the behavior of the __GDP__ using the three equivalent ways to measure production

### GDP as the sum of value added for Colombia using the ISIC [^1]

#### Sectors




[^1]: International Standard Industrial Classification of All Economic Activities

# Bibliography