---
title: "Economic concepts"
author: "Luis Francisco Gomez Lopez"
date: 2020-07-13
output:
  html_document:
    code_folding: hide
    df_print: paged
    highlight: tango
    number_sections: yes
    theme: flatly
    toc: yes
    toc_depth: 3
bibliography: economic_concepts.bib
link-citations: yes
---

<style>

body {
text-align: justify}

</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo      = TRUE, 
                      warning   = FALSE, 
                      message   = FALSE,
                      fig.align = "center")
```

```{r libraries}

library(tidyverse)
library(wbstats)
library(plotly)
```

# Gross Domestic Product (GDP)

We can initially approach the concept of __GDP__ by explaining the terms __"Product"__, __"Domestic"__ and __"Gross"__ separately [@lequiller_understanding_2014, Chapter 1]:

- __Product:__ refers to what we are trying to measure, which is the production of goods and services, with no double counting, in a given period, $t$ and carried out by:

    + Profit-making enterprises (economists use the alternative term _firms_)
    
    + Non-profit institutions
    
    + Goverment bodies
    
    + Households

- __"Domestic"__: indicates that the production to be taken into account is the one that is carried within a certain geographical territory, $s$, clearly delimited. 

    + In that sense if we use the concept of of __GDP__ we don't care who carried out the production in $t$ but only where the production was carried out in $t$ 
    
- __"Gross"__: it means that depreciation or the cost of using capital is not deducted (in the field of economics it is called consumption of fixed capital). 

    + In other words, the decrease in the value of the assets due to physical deterioration, foreseeable wear or accidental damage is not deducted.
    
    + Why depreciation or the cost of using capital is not deducted?
    
        + It is a variable difficult to measure and the methods used to measure it tend to differ between countries creating problems for international comparison
        
## Three equivalent ways to measure the GDP

Initially the units  in which __GDP__ is measure is in monetary units of a specific currency, $c$. Therefore $GDP_{s}^{c}(t)$ means the $GDP$ of territory $s$ in a given period $t$. To make the discussion less abstract we present a plot of $GDP$ for Colombia, $s = COL$, expressed in Colombian pesos, $c = COP$, for the years 1960 to 2019, $t = 1960, \ldots, 2019$: 

```{r, fig.width = 7, fig.height = 8}

# Clean data
gdp_colombia <- wbstats::wb(country   = "COL", 
                             indicator = "NY.GDP.MKTP.CN", 
                             startdate = 1960, 
                             enddate   = 2019) %>% 
    tibble::as_tibble() %>% 
    dplyr::select(date, value) %>%
    dplyr::mutate(date       = as.double(date),
                  label_text = stringr::str_glue('Year: {date}
                                                  GDP: {value %>% scales::dollar()}')) 

# Plot
    
static_plot <- gdp_colombia %>%
    # Data
    ggplot2::ggplot(aes(x = date, y =value)) +
    # Geoms
    ggplot2::geom_point(aes(text = label_text),
                        shape = 21, 
                        color = "black", 
                        fill  = "red") + 
    ggplot2::geom_line(linetype = "dashed") +
    # Scales
    scale_x_continuous(breaks = c(1960:2019)) +
    scale_y_continuous(breaks = seq(from = 0, to = 1.10e15, by = 1e14), 
                       labels = scales::number_format(scale = 1/1e12, suffix = "B")) +
    labs(x = "Year",
         y = "GDP in current local currency [B = Billion in long scale (10^12)]",
         title = "GDP of Colombia: 1960-2019") +
    # Themes
    theme(panel.border      = element_rect(fill = NA, color = "black"),
          plot.background   = element_rect(fill = "#f3fcfc"),          
          panel.background  = element_rect(fill = "#f3f7fc"),
          legend.background = element_rect(fill = "#f3fcfc"), 
          plot.title        = element_text(face = "bold"),
          axis.text.x=element_text(angle = -90, vjust = 0.5),
          axis.title        = element_text(face = "bold"),
          legend.title      = element_text(face = "bold"),
          axis.text         = element_text(face = "bold"))

    # Interactivity
    static_plot %>% 
      plotly::ggplotly(tooltip = "text")
    
```

# Bibliography