---
title: Some simple approaches to improve the Welch- Satterthwaite approximation
author:
  - name: Carlos Alberto Cardozo Delgado
    acknowledgements: | 
      The authors gratefully acknowledge _please remember to list all relevant funding sources in the non-anonymized (unblinded) version_.
    affiliations: 
        - name: Pontificia Universidad Javeriana
          department: Departamento de Matem치ticas
  - name: Luis Francisco G칩mez L칩pez
    affiliations: 
        - name: Universidad Militar Nueva Granada
          department: Administraci칩n de Empresas
abstract: |
  In this work we propose three simple ways to improve the classical Welch-Satterthwaite (WS) approximation to the effective degree of freedom of a non-negative linear combination of $\chi^2$ distributions. The WS option is typically used in design of experiments and metrology. However, it has been pointing out in many references the multiple limitations of the inferences based on the WS approximation. Three novel  estimators of the effective degree of freedom of a non-negative linear combination of $\chi^2$ distributions are given. We also study some theoretical properties of the proposed estimators. Additionally, through several Monte Carlo simulations, we assess the bias, variance, and mean square error of the proposed estimators under (very) small and moderate sample sizes. The proposed estimators have a much better performance than the WS proposal  and his implications. Finally, two applications are presented in which the proposed estimators help to improve the performance of some interval estimation and hypothesis testing procedures.
keywords: 
  - Asymmetric distributions
  - Unbiased estimation
  - Consistent estimation
  - Monte Carlo simulation
date: last-modified
bibliography: bibliography.bib
format:
  jasa-pdf:
    keep-tex: true  
    journal:
      blinded: false
  jasa-html: default
---

```{r}
#| echo: false
#| message: false
library(tidyverse)
library(kableExtra)
library(latex2exp)
library(tidyquant)
```

## Introduction {#sec-intro}


## Critical aspects of the WS approach

@lloyd_2-sample_2013 conducted three numerical experiments highlighting potential limitations of the two-sample t-test when using the Welch-Satterthwaite (WS) approximation. Notably, in their first experiment, @lloyd_2-sample_2013 assumed that the ratio of population variances was known. However, in practice, this information is rarely available. Therefore, we will replicate this simulation without assuming a known variance ratio, providing a more realistic assessment of the two-sample t-test's performance under the WS approximation.

We now outline the initial simulation plan associated with the two-sample t-test statistic, $T = \frac{\overline{X} - \overline{Y} - (\mu_1 - \mu_2)}{\sqrt{\frac{\widehat{\sigma}^2_1}{n_1} + \frac{\widehat{\sigma}^2_2}{n_2}}}$. Let $X_1, X_2, \ldots, X_{n_1} \overset{\text{iid}}{\sim} \mathcal{N}(\mu_1, \sigma_1^2)$ and $Y_1, Y_2, \ldots, Y_{n_2} \overset{\text{iid}}{\sim} \mathcal{N}(\mu_2, \sigma_2^2)$ where $\mathbf{X} = (X_1, X_2, \ldots, X_{n_1})$ and $\mathbf{Y} = (Y_1, Y_2, \ldots, Y_{n_2})$ are 2 independent random vectors. We assume the following parameters:

-   $\mu_1 = 1$ and $\mu_2 = 2$.
-   $\sigma^2_1 = 1$ and varying variance ratios, $\frac{\sigma^2_2}{\sigma^2_1} \equiv \rho = 2, 4, 6, 8, 10, 12$.
-   $n_1$ takes values $n_1 = 2, 3, \ldots, 11$, with sample size ratios, $\frac{n_2}{n_1} \equiv k = 1, 2, 4$.

For each combination of $(\rho, n_1, k)$, we construct the sampling distribution of the two-sample t-test statistic based on $R = 10000$ replications. In contrast to @lloyd_2-sample_2013, we do not pre-calculate and fix the degrees of freedom, $\nu$, using the WS approach. Instead, for each replication, we estimate the degrees of freedom, denoted by $\widehat{\nu}$, using the WS approach (see @sec-df-ws for details). Next, we determine the lower and upper critical values, $t_{\frac{\alpha}{2},\widehat{\nu}}$ and  $t_{1-\frac{\alpha}{2},\widehat{\nu}}$, respectively for $\alpha = 0.05, 0.10$ to calculate the corresponding type I error rates for each combination of $(\rho, n_1, k)$ based on the WS approach as: 

$$\widehat{\alpha}_r = F_{T}\left(t_{\frac{\alpha}{2},\widehat{\nu}_r}\right) + \left(1 - F_{T}\left(t_{1-\frac{\alpha}{2},\widehat{\nu}_r}\right)\right) \text{ for } r = 1, \ldots, R$$ {#eq-type-I-error-rates}

where $F_T$ represents the empirical cumulative distribution function of the two-sample t-test statistic.

Finally, to assess the accuracy of the estimated type I error rates, we calculate the relative difference between each $\widehat{\alpha}_r$  and the corresponding true significance level $\alpha$ and compare it to a threshold of $d=0.10$:

$$\frac{|\widehat{\alpha}_r - \alpha|}{\alpha} > d = 0.10$$ {#eq-pro-rel-alpha}

@tbl-sim-pro-rel-alpha-005 and @tbl-sim-pro-rel-alpha-010 present the proportion of replicates satisfying @eq-pro-rel-alpha for the combination $(\rho, n_1, k)$ under both $\alpha = 0.05$ and $\alpha = 0.10$. Also @fig-sim-pro-rel-alpha visually represents these results, demonstrating that the accuracy of the estimated type I error rates generally improves as  $n_1$ increases and $k$ decreases. However, the effect of the variance ratio, $\rho$, on accuracy is not as clear-cut and warrants further investigation.

```{r}
#| echo: false
#| message: false
#| label: tbl-sim-pro-rel-alpha-005
#| tbl-cap: Proportion of replicates where the relative absolute difference between a nominal $\alpha$ and the $\alpha$-value given by WS approximation test is bigger than $d=0.10$ where $\alpha = 0.05$

simulation_pct_rel_dist_tbl <- read_csv(file = "data/simulation_pct_rel_dist_tbl.csv")

data_table <- simulation_pct_rel_dist_tbl |> 
  pivot_wider(id_cols = c(rho, alpha, k), 
              names_from = n_1, 
              values_from = pct_greater_d) |> 
  arrange(alpha, k, rho) |> 
  relocate(c(alpha, k), .before = rho)

data_table |>
  filter(alpha == 0.05) |> 
  mutate(across(.cols = `2`:`11`,
                .fns = \(x) round(x = x, digits = 3))) |>
  kbl(col.names = c("$\\alpha$",
                    "$k$",
                    "$\\rho$",
                    2:11),
      booktabs = TRUE,
      escape = FALSE,
      longtable = TRUE) |> 
  kable_styling(font_size = 7,
                latex_options = c("repeat_header",
                                  "hold_position")) |> 
  collapse_rows(columns = 1:2) |>  
  add_header_above(c(" " = 3,
                     "$n_1$" = 10), 
                   escape = FALSE)
```

```{r}
#| echo: false
#| message: false
#| label: tbl-sim-pro-rel-alpha-010
#| tbl-cap: Proportion of replicates where the relative absolute difference between a nominal $\alpha$ and the $\alpha$-value given by WS approximation test is bigger than $d=0.10$ where $\alpha = 0.10$

data_table |>
  filter(alpha == 0.10) |> 
  mutate(across(.cols = `2`:`11`,
                .fns = \(x) round(x = x, digits = 3))) |>
  kbl(col.names = c("$\\alpha$",
                    "$k$",
                    "$\\rho$",
                    2:11),
      booktabs = TRUE,
      escape = FALSE,
      longtable = TRUE) |> 
  kable_styling(font_size = 7,
                latex_options = c("repeat_header",
                                  "hold_position")) |> 
  collapse_rows(columns = 1:2) |>  
  add_header_above(c(" " = 3,
                     "$n_1$" = 10), 
                   escape = FALSE)
```

```{r}
#| echo: false
#| message: false
#| fig-align: center
#| fig-height: 7
#| fig-width: 7
#| label: fig-sim-pro-rel-alpha
#| fig-cap: Graphical representation of @tbl-sim-pro-rel-alpha-005 and @tbl-sim-pro-rel-alpha-010
#| fig-pos: h

figure_table1 <- simulation_pct_rel_dist_tbl |> 
  mutate(across(.cols = c(k, rho, alpha), 
                .fns = as.factor))

levels(figure_table1$k) <- c(TeX(r'($k=1$)'), 
                             TeX(r'($k=2$)'), 
                             TeX(r'($k=4$)'))
levels(figure_table1$alpha) <- c(TeX(r'($\alpha=0.05$)'), 
                                 TeX(r'($\alpha=0.10$)'))

figure_table1 |> 
  ggplot(aes(x = n_1, y = pct_greater_d)) + 
  geom_line(aes(color = factor(x = rho)),
            linewidth = 0.5) + 
  facet_wrap(facets = vars(k, alpha),
             nrow = 3,
             labeller = label_parsed) + 
  scale_color_tq() +
  scale_x_continuous(breaks = 2:11) +
  labs(x = TeX(r'($n_1$)'),
       y = TeX(r'(Proportion of replicates where $\frac{|\widehat{\alpha}_j - \alpha|}{\alpha} > d = 0.1$ is true)'),
       color = TeX(r'($\rho$)'))
```

## Conclusion {#sec-conc}


## Appendix {.appendix}

### Degrees of freedom estimated by the WS approach {#sec-df-ws}

In our case, following @satterthwaite_approximate_1946 and @welch_generalization_1947, the degrees of freedom $\widehat{\nu}$  can be expressed as:

$$\begin{split}
  \widehat{\nu} & = \frac{\left( \frac{\widehat{\sigma}^2_1}{n_1} + \frac{\widehat{\sigma}^2_2}{n_2} \right)^2}{\frac{1}{n_1-1} \left( \frac{\widehat{\sigma}^2_1}{n_1} \right)^2 + \frac{1}{n_2-1} \left( \frac{\widehat{\sigma}^2_2}{n_2} \right)^2} \\
  & = \frac{\left( \frac{\widehat{\sigma}^2_1}{n_1} + \frac{\widehat{\rho}\widehat{\sigma}^2_1}{kn_1} \right)^2}{\frac{1}{n_1-1} \left( \frac{\widehat{\sigma}^2_1}{n_1} \right)^2 + \frac{1}{kn_1-1} \left( \frac{\widehat{\rho}\widehat{\sigma}^2_1}{kn_1} \right)^2} \text{ where } \frac{\widehat{\sigma}^2_2}{\widehat{\sigma}^2_1} \equiv \widehat{\rho} \\
  & = \frac{(n_1-1)(kn_1-1)(k+\widehat{\rho})^2}{(kn_1-1)k^2 + (n_1-1)\widehat{\rho}^2}
  \end{split}$$ {#eq-df-ws}


## Supplementary Material {.supplementary}

functions.R

:   R script that contains the functions used to generate the different results

script_simulations_tables_1_2.R

:   R script used to produce the simulation results presented in @tbl-sim-pro-rel-alpha-005, @tbl-sim-pro-rel-alpha-010 and @fig-sim-pro-rel-alpha

simulation_pct_rel_dist_tbl.csv

:   Simulated dataset @tbl-sim-pro-rel-alpha-005 and @tbl-sim-pro-rel-alpha-010  

## References {-}
